---
title: "Getting started with freqseas"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with freqseas}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6.5,
  fig.height = 4.2
)
```

freqseas organizes seasonal adjustment around the workflow's own epistemic
hierarchy, rather than treating "adjust the series" as one undifferentiated
step:

* **detect** -- is there excess mass at seasonal frequencies at all, against a
  white-noise null? (testable)
* **specify** -- is the seasonality a line (a fixed, deterministic pattern) or
  a band (a stochastic process concentrated near the seasonal frequencies)?
  (testable)
* **identify** -- does the seasonality distort the series' own dynamics
  (the "filter" world) or sit on top of it as an independent, additive
  component (the "components" world)? This is a modelling commitment, not
  something the data can decide on their own -- see `vignette("two-worlds")`.
* **operate** -- given the above, remove the seasonal filter (gain surgery)
  and recolor the result back onto the original series.

This vignette walks the one-call path, then the equivalent two-stage path,
on `UKgas`, a small quarterly series built into base R (`datasets::UKgas`,
UK gas consumption, 1960 Q1 -- 1986 Q4). No extra data downloads, no random
seeds to worry about -- everything below is real, executed output.

```{r setup}
library(freqseas)
UKgas
```

## The one-call path

`seas_adjust()` runs the whole workflow -- test, specify, identify, operate --
in one call. `phase_rule` defaults to `"minimum"` at this wrapper layer (the
filter-world / deconvolution rule); see `vignette("two-worlds")` for why that
default exists and when `"zero"` is the right choice instead.

```{r one-call}
fit <- seas_adjust(UKgas)
fit
```

The printed summary already answers the workflow's first three questions for
this series: `UKgas` **is** seasonal (detection p far below 0.05), the
seasonality is a **band** (not a fixed line), and the declared identification
is tagged `[declared identification]` -- a reminder that "minimum" is a
choice, not a fact the data proved. The `comb:` line, when present, is a
diagnostic explained in `vignette("two-worlds")`; it flags when the excess
seasonal-frequency mass is concentrated at only *some* of the harmonics
(here, the annual one) rather than the full comb, which can also happen for
genuine, non-comb-shaped seasonality and is worth a second look rather than
an alarm.

`summary()` gives the full record behind that compact print: the detection
statistic and ordinate counts, the whitener's differencing/AR decision and
its BIC path, the bin-count selection, and the gain surgery actually applied.

```{r summary}
summary(fit)
```

### Plots

`plot()` on a `seas_sa` defaults to the original-vs-adjusted series with the
removed seasonal component beneath it; `type = "periodogram"` shows the
standardized periodogram before and after adjustment side by side, so you can
see the seasonal-bin ordinates (blue) collapse toward the donor (nonseasonal)
level.

```{r plot-series}
plot(fit)
```

```{r plot-periodogram}
plot(fit, type = "periodogram")
```

### Accessors

Four small accessor generics pull the pieces back out without reaching into
list internals: `adjusted()`, `seasonal()`, `whitener()`, `decision()`.

```{r accessors}
head(adjusted(fit))
head(seasonal(fit))
decision(fit)
str(whitener(fit)[c("d", "d_reason", "p", "ar")])
```

### broom generics

`tidy()`, `glance()`, and `augment()` (from the `generics` package) give
tibble-shaped views for anyone used to the broom ecosystem: `glance()` is a
one-row model summary, `augment()` is observation-level (original, adjusted,
seasonal), and `tidy()` renders the per-harmonic diagnostic table when one is
available.

```{r broom}
glance(fit)
head(augment(fit))
```

## The two-stage path

`seas_adjust()` is exactly `seas_ssi(seas_test(x, ...), phase_rule = ...)` --
one code path, not two implementations that could drift apart. The two-stage
form is what you reach for when you want to inspect the test (or reuse it
under more than one `phase_rule`) before committing to an adjustment.

```{r two-stage}
tst <- seas_test(UKgas)
tst
```

```{r two-stage-plot}
plot(tst)
```

`seas_ssi()` takes that test and a *required* `phase_rule` -- there is
deliberately no default here, only at the `seas_adjust()` wrapper layer (see
`vignette("two-worlds")` for why omitting it raises a teaching error instead
of silently picking one):

```{r two-stage-ssi}
adj <- seas_ssi(tst, phase_rule = "minimum")
identical(adjusted(adj), adjusted(fit))
```

## A keyed-tsibble teaser

freqseas maps the whole workflow over the keys of a keyed `tsibble`
(fable-style): one fit per key, returned as a `seas_collection`. The example
below is a small **synthetic** two-key panel built purely to show the shape
of the API -- the full treatment (decision persistence across vintages, a
`vintagesim` round trip, `benchmark_totals()`) is in
`vignette("panels-and-vintages")`.

```{r panel-teaser}
library(tsibble)

sim_seasonal_q <- function(n, seed, Phi, phi = 0.5, s = 4L, spin = 300L) {
  set.seed(seed)
  ntot <- spin + n
  eps  <- stats::rnorm(ntot)
  z    <- numeric(ntot)
  for (t in (s + 2L):ntot) {
    z[t] <- phi * z[t - 1L] + Phi * z[t - s] - phi * Phi * z[t - s - 1L] + eps[t]
  }
  z[(spin + 1L):ntot]
}

panel <- tsibble::tsibble(
  region = rep(c("North", "South"), each = 80L),
  qtr    = rep(tsibble::yearquarter("2000 Q1") + 0:79, 2),
  value  = c(sim_seasonal_q(80L, seed = 1L, Phi = 0.9),
             sim_seasonal_q(80L, seed = 2L, Phi = 0.7)),
  key = region, index = qtr
)

coll <- seas_adjust(panel, phase_rule = "minimum")
coll
```

`adjusted()`/`seasonal()`/`whitener()`/`decision()` all have `seas_collection`
methods too, returning a tibble keyed the same way:

```{r panel-accessors}
decision(coll)
```

## Where next

* `vignette("two-worlds")` -- the identification story: why `phase_rule`
  can't be chosen by the data alone, a miniature verified demonstration, and
  the comb diagnostic seen above.
* `vignette("the-machinery")` -- bins, donors, EVT, the M0-Whittle whitener,
  `suggest_M()`, and `min_phase()` for readers who want the mechanics behind
  each stage.
* `vignette("panels-and-vintages")` -- keyed tsibbles end to end, decision
  persistence under concurrent adjustment, a `vintagesim` round trip, and
  `benchmark_totals()`.
