---
title: "Panels and vintages"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Panels and vintages}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6.5,
  fig.height = 4.2
)
```

```{r setup}
library(freqseas)
library(tsibble)
library(dplyr)
```

Two things distinguish practical, production seasonal adjustment from the
single-series demonstrations in the other vignettes: there are usually many
series at once (a **panel**), and each series keeps growing as new data
arrive (**vintages**). This vignette covers both, plus the NIPA-style
constraint that annual totals shouldn't move just because a series was
adjusted.

## Keyed tsibbles end to end

freqseas maps its whole workflow over the keys of a keyed `tsibble`
(fable-style): call `seas_test()`/`seas_ssi()`/`seas_adjust()` once on the
panel, get one fit per key back as a `seas_collection`. Each key is
adjusted independently -- there is no cross-series pooling (that's future
work) -- but the bookkeeping (key ordering, re-clothing each key's output
back into a tsibble, one row per key on print) is handled for you.

```{r panel-build}
sim_seasonal_q <- function(n, seed, Phi, phi = 0.5, s = 4L, spin = 300L) {
  set.seed(seed)
  ntot <- spin + n
  eps  <- stats::rnorm(ntot)
  z    <- numeric(ntot)
  for (t in (s + 2L):ntot) {
    z[t] <- phi * z[t - 1L] + Phi * z[t - s] - phi * Phi * z[t - s - 1L] + eps[t]
  }
  z[(spin + 1L):ntot]
}

panel <- tsibble::tsibble(
  region = rep(c("North", "South", "East"), each = 80L),
  qtr    = rep(tsibble::yearquarter("2000 Q1") + 0:79, 3),
  value  = c(sim_seasonal_q(80L, seed = 1L, Phi = 0.90),
             sim_seasonal_q(80L, seed = 2L, Phi = 0.70),
             sim_seasonal_q(80L, seed = 3L, Phi = 0.85)),
  key = region, index = qtr
)
panel
```

```{r panel-test}
tst_coll <- seas_test(panel)
tst_coll
```

`seas_ssi()` (or the one-call `seas_adjust()`) maps over the same keys;
`phase_rule` is required here exactly as it is for a single series:

```{r panel-adjust}
coll <- seas_ssi(tst_coll, phase_rule = "minimum")
coll
```

Every accessor and every generic works the same way, returning a tibble
keyed by the panel's own key columns instead of a single value:

```{r panel-generics}
decision(coll)
glance(coll)
head(augment(coll))
```

`adjusted()`/`seasonal()` re-clothe each key's result back into a `tbl_ts`,
so `adjusted(coll)$adjusted[[1]]` is itself a tsibble, not a bare numeric
vector:

```{r panel-reclothe}
adj_tbl <- adjusted(coll)
class(adj_tbl$adjusted[[1]])
```

## Decision persistence under concurrent adjustment

A vintage series doesn't arrive all at once -- it grows, one release at a
time, and each release is a longer prefix of (eventually) the same
underlying path. The tempting default is to re-run the whole
detect-specify-identify-operate workflow fresh on every release. That is
usually the wrong instinct: because the detection test operates at a fixed
significance level, a series whose seasonal signal sits anywhere near the
threshold will cross it back and forth as noise accumulates, and if you
retest every vintage, the *adjustment itself* silently switches on and off
along with it -- an artificial discontinuity that has nothing to do with new
information about the series and everything to do with resampling the same
test at a moving sample size.

This is not a hypothetical. Simulate one long path (a single realization,
the same way a real series has exactly one realized history) and take
growing prefixes of it as successive quarterly vintages:

```{r vintage-setup}
set.seed(19)
n_max <- 200L; Phi <- 0.35    # a modest seasonal signal, deliberately near
                               # the detection threshold at these lengths
e <- stats::rnorm(n_max + 400)
z <- numeric(n_max + 400)
for (t in 6:(n_max + 400)) {
  z[t] <- 0.5 * z[t - 1] + Phi * z[t - 4] - 0.5 * Phi * z[t - 5] + e[t]
}
full <- z[401:(n_max + 400)]
vintage_lens <- seq(80L, 160L, by = 8L)   # 11 successive quarterly releases
```

Re-testing fresh at every release:

```{r vintage-naive}
naive_decision <- vapply(vintage_lens, function(L) {
  seas_test(stats::ts(full[1:L], frequency = 4))$decision
}, logical(1))
data.frame(vintage = seq_along(vintage_lens), n = vintage_lens,
           decision = naive_decision)
```

The adjustment would turn on and off `r sum(diff(naive_decision) != 0)` times
across these 11 releases -- and each flip is a real, visible discontinuity:
the *same* historical quarter is reported seasonally adjusted in one release
and unadjusted (raw) in the very next, purely because one more quarter of
noise pushed a p-value across 0.05.

The alternative: re-decide only at scheduled reviews (here, every 5th
release -- an "annual review" cadence relative to the quarterly vintages),
and hold that decision fixed for every release in between:

```{r vintage-disciplined}
review_every <- 5L
disciplined_decision <- logical(length(vintage_lens))
current <- NA
for (i in seq_along(vintage_lens)) {
  if ((i - 1L) %% review_every == 0L) {
    current <- seas_test(stats::ts(full[1:vintage_lens[i]], frequency = 4))$decision
  }
  disciplined_decision[i] <- current
}
data.frame(vintage = seq_along(vintage_lens), n = vintage_lens,
           decision = disciplined_decision,
           is_review_point = (seq_along(vintage_lens) - 1L) %% review_every == 0L)
```

Review-gated testing asks the question 3 times instead of 11, and each
review reaches a stable answer that then governs every release until the
next one -- exactly the discipline the design notes call for: *"test/spec
decisions persist across vintages; re-decide at annual reviews only."* The
same logic extends past the on/off decision: once a review fixes `M`, the
line/band specification, and `phase_rule`, pass those explicitly (rather
than `M = NULL`, letting them float) to every `seas_adjust()` call before the
next review, so the *character* of the adjustment -- not just whether one
happens -- stays stable between reviews too.

## `benchmark_totals()`: annual totals are invariant to adjustment

A separate, NIPA-style constraint: seasonal adjustment should redistribute
activity *within* a year, not change the year's total. `benchmark_totals()`
proportionally rescales an adjusted series so each complete period's total
matches the original's -- one multiplicative factor per period, applied
within that period only.

```{r benchmark}
data(ssi_examples)
orig <- ssi_examples$x[[which(ssi_examples$world == "filter")]]
adj  <- seas_adjust(orig, phase_rule = "minimum")
matched <- benchmark_totals(adjusted(adj), orig)

N  <- stats::frequency(orig)
yo <- colSums(matrix(as.numeric(orig),             nrow = N))  # original annual totals
ya <- colSums(matrix(as.numeric(adjusted(adj)),     nrow = N))  # adjusted, unmatched
ym <- colSums(matrix(as.numeric(matched),           nrow = N))  # adjusted, benchmarked

max(abs(ya - yo))   # unmatched: adjustment moved the annual totals
max(abs(ym - yo))   # benchmarked: annual totals restored to ~0
```

## A `vintagesim` round trip

The `vintagesim` package (a Suggests dependency, used here and in the
package's own integration tests) simulates a small vintage environment with
a known latent nonseasonal truth, letting an adjustment be checked against a
ground truth no real dataset provides. The chunk below only runs if
`vintagesim` is installed.

```{r vintagesim-check, echo = FALSE}
have_vintagesim <- requireNamespace("vintagesim", quietly = TRUE)
```

```{r vintagesim-round-trip, eval = have_vintagesim}
spec <- vintagesim::new_simulation_spec(
  J = 4, sample_years = 4, history_years = 6, seed = 101
)
env  <- vintagesim::simulate_vintage_environment(spec)
v    <- vintagesim::vintage_range(spec)
snap <- vintagesim::get_snapshot(env, v[length(v)])

keys <- c("C01", "C02", "C03")
tb <- tibble::as_tibble(snap)
tb <- dplyr::filter(tb, process == "seasonal", series_type == "component",
                    series_id %in% keys)
tb <- dplyr::select(tb, series_id, period, value)
tb <- tsibble::as_tsibble(tb, key = series_id, index = period)

vs_coll <- seas_adjust(tb, phase_rule = "minimum")
vs_coll
```

The latent nonseasonal truth lives in `env$truth`, keyed the same way as the
snapshot; comparing the adjusted series against it (as
`tests/testthat/test-integration-vintagesim.R` does across many seeds) is
how the package's own tier-4 acceptance test confirms that adjustment beats
no adjustment against a ground truth outside of a purely synthetic
`ssi_examples` draw. One component here, compared demeaned (the shared-
innovation filter's frequency-zero gain means levels aren't recoverable --
the same fact `vignette("two-worlds")` establishes):

```{r vintagesim-compare, eval = have_vintagesim}
nonseasonal_truth <- function(env, sid) {
  df <- tibble::as_tibble(env$truth)
  df <- dplyr::filter(df, process == "nonseasonal", series_type == "component",
                      series_id == sid)
  dplyr::arrange(df, period)
}

sid <- "C01"
truth_df <- nonseasonal_truth(env, sid)
obs_df   <- dplyr::arrange(dplyr::filter(tb, series_id == sid), period)
common   <- intersect(truth_df$period, obs_df$period)

xv  <- obs_df$value[match(common, obs_df$period)]
txv <- truth_df$value[match(common, truth_df$period)]

adj_tbl_vs <- adjusted(vs_coll)
adj_ts_vs  <- adj_tbl_vs$adjusted[[which(adj_tbl_vs$series_id == sid)]]
av <- as.numeric(adj_ts_vs[[tsibble::measured_vars(adj_ts_vs)]])

dm <- function(z) z - mean(z)
c(none = sqrt(mean((dm(xv) - dm(txv))^2)),
  adj  = sqrt(mean((dm(av) - dm(txv))^2)))
```

```{r vintagesim-missing, eval = !have_vintagesim, echo = FALSE, results = "asis"}
cat("_`vintagesim` is not installed in the environment that built this vignette;",
    "the round-trip chunks above were skipped._\n")
```

## Where next

* `vignette("getting-started")` -- the same accessors and generics on a
  single series, without the panel/vintage layer.
* `vignette("two-worlds")` -- why `phase_rule` has to be declared once
  (ideally at the same review cadence as the rest of the decision).
* `vignette("the-machinery")` -- what actually changes inside the adjustment
  when `M`, `d`, or the AR order drift between vintages.
