---
title: "The machinery: bins, donors, whitening, and surgery"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The machinery: bins, donors, whitening, and surgery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6.5,
  fig.height = 4.2
)
```

```{r setup}
library(freqseas)
```

`vignette("getting-started")` shows *what* to call and `vignette("two-worlds")`
explains *which identification* to declare. This vignette is for readers who
want to see *how* each stage works -- the bin geometry, the detection
statistic, the whitener, the phase reconstruction, and the surgery itself --
matched to the paper's notation. The internal functions that implement each
stage (`fs_whiten()`, `fs_donor_pool()`, `fs_gain()`, `fs_surgery()`, and the
bin/index-set builders) are not exported -- they are implementation, not
API -- so everything below is shown through the fields a `seas_test`/`seas_sa`
object actually returns, which *are* part of the documented contract.

One worked example runs through the whole vignette: a quarterly
AR(2)$\times$SAR(1)-type series, $n = 120$, seasonal coefficient $\Phi = 0.8$
(the same generator used in most of the package's own `@examples`):

```{r worked-example}
set.seed(1)
n <- 120L; Phi <- 0.8
e <- stats::rnorm(n + 400)
z <- numeric(n + 400)
for (t in 6:(n + 400)) {
  z[t] <- 0.5 * z[t - 1] + Phi * z[t - 4] - 0.5 * Phi * z[t - 5] + e[t]
}
x <- stats::ts(z[401:(n + 400)], frequency = 4)
tst <- seas_test(x)
```

## Bins: the seasonal/nonseasonal partition

The first geometric object in the pipeline is a partition of the positive
frequency axis $(0, \pi]$ into $M$ equal-width bins. A bin is **seasonal**
($M_1$) if it contains one of the seasonal frequencies
$\Omega_G = \{2\pi k / (PN) : k = 1, \dots, \lfloor PN/2 \rfloor\}$ (for
quarterly data with the fundamental multiplier $P = 1$, that's just $\pi/2$
and $\pi$); every other bin is **nonseasonal** ($M_0$). `tst$partition`
carries the realized partition:

```{r partition}
tst$partition$breaks              # bin edges on (0, pi]
tst$partition$M1                  # seasonal bin indices
tst$partition$M0                  # nonseasonal bin indices
```

That bin classification is then mapped onto actual Fourier-ordinate indices
of the whitened series (`J1`/`J0`, `N1`/`N0` seasonal/nonseasonal ordinate
counts, `H` the exact-harmonic ordinates, `S` the shoulder ordinates
`J1 \ H`):

```{r index-sets}
tst$partition$N1                  # seasonal-bin ordinate count
tst$partition$N0                  # nonseasonal-bin (donor) ordinate count
```

## Donors and the EVT detection test

**Detection** asks a single question: is there more power at the seasonal
ordinates ($J_1$) than a white-noise process would produce, relative to the
nonseasonal ordinates ($J_0$) as a reference? The statistic is a max-contrast,
$\Delta = \max_{J_1} I - \max_{J_0} I$, on the *standardized* periodogram of
the whitened residuals. Under the white null this statistic is (to a good
approximation) logistic with location $\hat\tau \log(N_1/N_0)$ and scale
$\hat\tau$ -- an extreme-value argument, not a chi-squared one, because it's a
contrast of *maxima*, and it is what gives the test its name (EVT).

```{r evt}
tst$evt
```

The **donor pool** is a separate, smaller idea: which nonseasonal ordinates
serve as the "clean" reference level when it comes time to estimate *gain*
(Stage C, below). It's the $J_0$ ordinates at or below the
`donor_quantile` (default 0.9) of the $J_0$ periodogram distribution --
excluding any nonseasonal ordinate that happens to carry unusually high
power (leakage from a nearby peak, or any other spectral feature) from
inflating the reference level:

```{r donors}
length(tst$donor_pool)            # size of the donor pool
tst$donor_level                   # reference (donor) periodogram level
```

## `suggest_M()`: choosing the bin count

$M$ trades off two things: too few bins and a seasonal bin swallows too much
of the nonseasonal spectrum next to it (contaminating the donor reference);
too many bins and each one holds too few Fourier ordinates to estimate
anything from. `suggest_M()` picks $M$ so that every seasonal frequency gets
its own bin (no merging), no seasonal frequency sits on a bin boundary, and
the seasonal-bin share $K_M/M$ lands near a target (default 0.25):

```{r suggest-m}
suggest_M(P = 1, N = 4, n_obs = 120)
```

freqseas's binning is **offset-free** (bins are the fixed partition
$((m-1)\pi/M,\, m\pi/M]$); a nonzero-offset partition can occasionally reach a
smaller $M$, but is not realizable by the package's binning:

```{r suggest-m-offset}
suggest_M(P = 1, N = 4, n_obs = 120, offset_grid = c(0, 0.5))
```

`seas_test()` uses this offset-free suggester by default (`M = NULL`), and
always reports how `M` was chosen:

```{r m-selection}
tst$M_selection
```

## The M0-Whittle whitener

Before any of the above runs, the series is **whitened**: a small AR filter
is fit and subtracted so that what's left (`e`, the whitened residuals) is as
close to white noise as a modest-order AR model can make it, without ever
being allowed to explain away the seasonality itself. The whitener's
blindness to seasonality is a design guarantee, not an accident of AR order:
at quarterly frequency an AR(2) with a negative lag-2 coefficient already
peaks at $\pi/2$, so simply *capping* the order at, say, 3 is not protection
-- what protects the fit is that its likelihood only ever *looks at*
nonseasonal ordinates. The default geometry (`exclusion = "guard"`) excludes
only the ordinates within a small guard band (default 3 Fourier steps) of
each exact seasonal harmonic, leaving the broad shoulders of a genuine
nonseasonal peak at a seasonal frequency visible to the fit -- which is what
lets the whitener explain such a peak (curing false alarms) while staying
blind to a concentrated seasonal comb.

The model fit to those included ordinates is not the naive infinite-sample AR
spectral density, but the **de-biased expected periodogram** of Sykulski,
Olhede, Guillaumin, Lilly & Early (2019, *Biometrika*) -- the finite-$n$
expectation $E[I(\omega_j)]$ of the periodogram under the candidate AR model,
computed exactly at the Fourier frequencies from the (Bartlett-weighted)
theoretical autocovariance. Fitting to this finite-sample expectation, rather
than to the $n \to \infty$ spectral density, removes the systematic
under-extrapolation of a peak's height that a bin-restricted fit otherwise
shows.

Order selection is a de-biased-Whittle BIC search over `p = 0..ar_max`
(default cap 3):

```{r whitener}
tst$whitener[c("d", "d_reason", "p", "ar", "sigma2")]
tst$whitener$bic_path
tst$whitener[c("exclusion", "guard")]
```

The guard geometry's default (`guard = 3`) was chosen by a fixed-seed
experiment contrasting it against whole-bin exclusion and a narrower guard
(see `?fs_whiten`'s Details for the full measured matrix); the key numbers,
at $n = 120$ quarterly: guard(3) cuts the false-alarm rate on a genuine
nonseasonal AR(2) cycle at $\pi/2$ ($\rho = 0.85$) from about 0.56 (whole-bin
exclusion) to about 0.11-0.15, while a comb-shaped seasonal signal's
bin-power ratio after whitening stays around 19 -- meaning a real comb
survives whitening essentially intact, which is the property the design
exists to guarantee. `vignette("two-worlds")` demonstrates the residual hard
case (a narrower cycle that still occasionally trips detection) and the
diagnostic (`comb_note`) that flags it after the fact.

## `min_phase()`: phase from gain alone

Once a seasonal band is detected, Stage C needs a *filter* to remove, not
just a gain -- and (per `vignette("two-worlds")`) a gain alone doesn't pin
down a filter's phase. `min_phase()` computes the phase of the unique
causal, causally-invertible filter sharing a given gain, via the discrete
Hilbert-transform relation between log-gain and phase, implemented as a
cepstral fold (Oppenheim & Schafer): take the real cepstrum of the log-gain,
double its causal (positive-lag) half onto itself, zero out the anticausal
half, and transform back -- the fold is exactly where causality enters.

A closed-form case makes this concrete: the seasonal-AR-type filter
$H(\omega) = 1/(1 - \Phi e^{-4i\omega})$ has gain
$G(\omega) = |H(\omega)|$ and a known phase, a sawtooth that rotates through
each seasonal harmonic:

```{r min-phase}
Phi <- 0.9; n <- 80L
w  <- 2 * pi * (0:(n - 1)) / n
G  <- 1 / sqrt(1 + Phi^2 - 2 * Phi * cos(4 * w))
theta_true <- -atan2(Phi * sin(4 * w), 1 - Phi * cos(4 * w))
theta_hat  <- min_phase(log(G))
max(abs(theta_hat - theta_true))
```

At only $n = 80$ knots for a fairly peaked gain ($\Phi = 0.9$), linear
interpolation to the fine internal grid under-resolves the peak's curvature,
and the error saturates around 0.02 rad regardless of how fine that internal
grid is set -- a resolution-boundary artifact of the $n = 80$ sampling, not
of the cepstral fold itself (which is exact to under $10^{-10}$ given an
accurate input; see `tests/testthat/test-minphase.R`). It shrinks as $n$
grows, and is already an order of magnitude below the phase errors that
matter for adjustment accuracy (`vignette("two-worlds")`'s RMSE comparisons).

## The surgery

Stage C removes the estimated filter from the DFT of the whitened residuals.
Two specifications:

* **band** -- every ordinate is divided by the estimated gain and de-rotated
  by the phase implied by the declared `phase_rule` (`min_phase(log(Ghat))`
  for `"minimum"`, zero for `"zero"`).
* **line** -- only the exact-harmonic ordinates are touched: their magnitude
  is reset to the donor level and their phase is left alone (a deterministic
  line has no filter phase to de-rotate).

Both branches enforce the same three guards: the DC ordinate is left
byte-for-byte identical to the input, the Nyquist ordinate (if present) is
forced exactly real, and Hermitian symmetry is rebuilt exactly (rather than
trusted to floating-point arithmetic) for every ordinate the branch touched.
These guards -- plus a taper/pad discipline against circular wraparound at
the sample edges -- are what keep the surgery numerically well-behaved at the
boundaries, which matters most exactly where vintage revisions matter most:
the most recent observations.

```{r surgery}
adj <- seas_ssi(tst, phase_rule = "minimum")
range(adj$Ghat)                   # gain applied (>= 1; 1 == untouched)
range(adj$theta)                  # phase applied (radians)
```

The gain itself is never estimated ordinate-by-ordinate raw: it's floored at
1 (the surgery never *amplifies* an ordinate relative to the donor level),
then lightly smoothed in log space across the seasonal bins so the
correction doesn't chase single-ordinate periodogram noise, and tapers
smoothly back to the donor level (gain 1) just outside the seasonal bins
rather than presenting a hard cliff at the bin edge.

## Recoloring

The adjusted innovations are not simply substituted for the original ones
and forecast forward from stale fitted values (`fitted + e*`) -- that
reintroduces exactly the seasonality Stage C just removed, through the AR
model's own memory of the un-adjusted history (this was the root cause
diagnosed in the design notes' original comparison run). Instead, recoloring
runs the *same* AR recursion that produced the whitened residuals, but
**forward through the adjusted innovations**, and undoes any differencing by
anchoring at the original series' own initial level:

```{r recolor}
max(abs(as.numeric(adj$adjusted) + as.numeric(adj$seasonal) - as.numeric(x)))
head(as.numeric(adj$adjusted), 3)   # first p + d values: held at x's own level
head(as.numeric(x), 3)
```

Seasonality can only re-enter the reconstruction through the innovations
themselves -- which is exactly what Stage C cleaned.

## Where next

* `vignette("two-worlds")` -- why `phase_rule` has to be declared, and what
  it costs to declare it wrong.
* `vignette("panels-and-vintages")` -- the same machinery mapped over keyed
  panels, and why test/spec decisions should persist across vintages rather
  than being re-run on every release.
