---
title: "Two worlds: filter versus components identification"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Two worlds: filter versus components identification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6.5,
  fig.height = 4.2
)
```

```{r setup}
library(freqseas)
```

This is the vignette [`seas_ssi()`](../reference/seas_ssi.html)'s error message
points to. Try to adjust a series without declaring `phase_rule` and you get a
teaching error, not a default:

```{r error-demo}
data(ssi_examples)
tst <- seas_test(ssi_examples$x[[1]])
tryCatch(seas_ssi(tst), error = function(e) cat(conditionMessage(e), "\n"))
```

That error is the whole subject of this vignette. `phase_rule` is not a
tuning parameter you can cross-validate away -- it is a declaration of *which
world generated your data*, and the two worlds are observationally
equivalent at second order. This page explains why, shows a miniature but
fully verified demonstration of what happens when you get it right and
wrong, and ends with the diagnostics that inform (without fully resolving)
the choice.

## Why gain alone doesn't determine a filter

Seasonal adjustment removes a filter: some linear operation
$H(\omega) = G(\omega) e^{i\theta(\omega)}$ that the observed series carries
relative to its nonseasonal counterfactual. A periodogram only ever tells you
about *power*, i.e. about $G(\omega) = |H(\omega)|$, the gain. It says nothing
about $\theta(\omega)$, the phase.

That is a real gap, not a technicality. Multiply $H$ by any all-pass factor
(magnitude 1 at every frequency, arbitrary phase) and the gain is unchanged
while the filter -- and the time-domain sequence it produces -- is not. So
"estimate the gain from the periodogram, divide it out, leave the phase
alone" (classic ordinate replacement) is already smuggling in a choice: it
implicitly assumes the *zero-phase* filter among all the filters sharing that
gain. That choice is invisible unless you say it out loud, which is exactly
what `phase_rule` forces.

freqseas resolves the gap with two *named* rules, not an infinite family of
possibilities:

* **`"minimum"`** -- the unique filter, among all filters sharing a gain,
  that is both causal and causally invertible. Its phase is pinned to the
  gain by a Hilbert-transform relation, $\theta(\omega) =
  -\mathcal{H}\{\log G\}(\omega)$ (the mechanics are in
  `vignette("the-machinery")`).
* **`"zero"`** -- $\theta(\omega) \equiv 0$. The classic published-SSI rule:
  divide the gain, touch nothing else.

## The two worlds

The choice of rule is not arbitrary once you commit to a generative story for
the seasonality -- it falls out of it.

**Filter world.** Seasonality distorts the series' *own* dynamics: the
observed series and its nonseasonal counterfactual share one innovation
stream, related by $x_t = H(B)\, x^{NS}_t$ for a causal filter $H$. Removing
seasonality is *deconvolution*: you want $H^{-1}$, which requires both the
gain and the phase. If $H$ itself is minimum-phase (the natural assumption
for something built from a causal AR-type mechanism), `phase_rule =
"minimum"` recovers it.

**Components world.** Seasonality is an independent, additive component with
its own innovations: $x_t = x^{NS}_t + s_t$, $s \perp x^{NS}$. Removing it is
*signal extraction*, not deconvolution. The optimal (Wiener) filter here is
$W(\omega) = f_{NS}(\omega) / (f_{NS}(\omega) + f_s(\omega))$, which is
**real** -- zero phase, by construction. There is no filter phase to
de-rotate because $x$'s phase at the seasonal frequencies is just $s$'s own
independent noise. Rotating it (`"minimum"`) doesn't recover anything; it
actively distorts a component that needed no rotation. `phase_rule = "zero"`
is correct here, and it is what classic published SSI has always
implicitly assumed.

**Both stories produce the same second-order spectrum.** Any spectrum
factors either as $G^2 f_{NS}$ (a filter-world reading) or as
$f_{NS} + f_s$ (a components-world reading with $f_s = (G^2 - 1) f_{NS}$) --
the two are spectrally indistinguishable. Under Gaussianity this is
undecidable *in principle* from the data alone: the two counterfactual paths
differ, but their distributions do not. That is why `seas_ssi()` refuses to
default `phase_rule` -- doing so would quietly assert an identification the
data cannot support. (A non-Gaussian escape hatch exists in principle --
bicoherence tests can, in theory, distinguish a single nonlinear-in-phase
stream from a sum of two independent linear streams -- but freqseas does not
currently implement it; see the architecture notes for the citation trail.)

## When each rule is right

| your belief about the DGP | rule | what "removal" means |
|---|---|---|
| seasonality is baked into the series' own filter (shared innovations) | `"minimum"` | deconvolution: divide gain, de-rotate by the implied minimum phase |
| seasonality is an independent additive component | `"zero"` | Wiener-style shrinkage: divide gain, leave phase alone |
| seasonality is a fixed, deterministic pattern (not a filter of anything) | *(either -- see below)* | reset the exact-harmonic ordinates to the donor level; no band phase to speak of |

## A verified demonstration

`ssi_examples` bundles one simulated quarterly series per world, **paired
with its true nonseasonal counterfactual** -- a comparison no real dataset
permits. Both worlds share the same nonseasonal AR(1) building block ($\phi =
0.5$) and the same seasonal strength ($\Phi = 0.9$); only the *mechanism*
connecting them to the observed series differs (see `?ssi_examples` for the
exact equations).

```{r worlds-setup}
dm   <- function(z) z - mean(z)
rmse <- function(a, b) sqrt(mean((dm(a) - dm(b))^2))

run_world <- function(world) {
  row   <- ssi_examples[ssi_examples$world == world, ]
  x     <- row$x[[1]]
  truth <- row$truth[[1]]
  amin  <- seas_adjust(x, phase_rule = "minimum")
  azero <- seas_adjust(x, phase_rule = "zero")
  data.frame(
    world = world,
    none    = rmse(as.numeric(x), truth),
    minimum = rmse(as.numeric(adjusted(amin)),  truth),
    zero    = rmse(as.numeric(adjusted(azero)), truth)
  )
}

result <- rbind(run_world("filter"), run_world("components"))
result
```

(RMSE is against the *demeaned* truth throughout this vignette, matching the
package's own test suite: the shared-innovation filter in the filter world
has the same gain at frequency zero as at the seasonal frequencies, so no
adjustment recovers the counterfactual's *level* -- only its shape around
its own mean. That is not a shortcoming of the adjustment; it is a fact
about what the filter-world model can identify at all.)

On this single bundled draw: in the **filter** world, `"minimum"` beats
`"zero"` by a clear margin, and both beat no adjustment. In the
**components** world, `"zero"` edges out `"minimum"` -- but the margin here
is thin (`r sprintf("%.4f", result$zero[result$world == "components"])` vs.
`r sprintf("%.4f", result$minimum[result$world == "components"])`), because
the gain has to be *estimated* from a single donor pool at $n = 120$, and the
two rules only differ in a phase correction that the components world's true
filter doesn't need much of relative to sampling noise. A single draw is not
a reliable estimate of a win rate. The package's test suite
(`tests/testthat/test-worlds.R`) reruns this exact comparison 30 times at
fixed seeds and checks the *robust* claim: `"minimum"` beats `"zero"` in the
filter world in at least 70% of replications (with a strict mean-RMSE
ordering), and `"zero"` beats `"minimum"` in the components world in at
least 60% of replications -- both directions holding with margin, and both
rules always beating no adjustment. That regression table, not any single
draw, is the claim to trust.

## The deterministic-line case

A third case sits outside both worlds: a genuinely **fixed**, non-random
seasonal pattern. It isn't a filter of anything (there's no innovation
stream driving it), so neither deconvolution nor Wiener shrinkage applies --
the correct surgery is simply to reset the exact-harmonic ordinates to the
donor (nonseasonal) level and leave phase alone (`spec$label == "line"`
routes to this branch regardless of `phase_rule`).

```{r line-default}
row <- ssi_examples[ssi_examples$world == "line", ]
x   <- row$x[[1]]
tst_default <- seas_test(x)
tst_default$spec[c("label", "shoulder_p", "phase_R")]
```

That the label comes back `"line"` is worth slowing down on, because the
thing standing behind it is a piece of grid bookkeeping that is easy to get
wrong. The shoulder test compares each exact seasonal harmonic against its
immediate neighbours, which means something only if those harmonics are
Fourier ordinates of the series actually analyzed -- the *whitened* residual
series of length `n_e`, not the original series of length `n`. Differencing
and an AR fit consume observations, so nothing about `n` being a multiple of
the seasonal period carries over to `n_e`. A deterministic line sitting
between ordinates leaks across the whole spectrum, the shoulders come up,
and the shoulder test reads band -- a fixed pattern misread as a stochastic
one.

`fs_whiten()` closes that gap by construction: it trims the residual vector
to a whole number of seasonal cycles before anything downstream sees it, so
`N` divides `n_e` always and the seasonal harmonics are always exact
ordinates. Here the whitener takes no difference and an AR(1) term, leaving
119 residuals, of which the last $116 = 29 \times 4$ are kept:

```{r line-trim}
tst_default$whitener[c("d", "p", "n_e", "n_trim")]
```

The cost is real, and bounded by `N - 1`. The trimmed residuals are spliced
back unmodified during recoloring, so the first `n_trim + d + p`
observations of an *adjusted* series are returned exactly as supplied --
four of 120 here. On monthly data that can be a full year;
`log(AirPassengers)` differences once and fits an AR(1), leaving 142
residuals to be trimmed to 132:

```{r trim-cost}
seas_test(log(AirPassengers))$whitener[c("d", "p", "n_e", "n_trim")]
```

so its first 12 observations -- 8% of the sample -- come back unadjusted.
That is not silent: `summary()` reports the trim in its whitener section
whenever it is nonzero. The trim is taken from the front deliberately. The
recent end of a series is the end that matters for seasonal adjustment, so
the edge effect is pushed onto the oldest observations, which is the
cheapest place to put it.

The block-phase resultant printed above is a separate, *descriptive*
corroborator rather than part of the shoulder test: `phase_R` close to 1
says the pattern is frozen rather than re-randomizing, and unlike the
shoulder test it never depended on the grid at all.

## The comb diagnostic

Even with the guard-based whitener (see `vignette("the-machinery")`), a
narrow nonseasonal cycle sitting *at* a seasonal frequency can still trip the
detection test -- it is not seasonality, but it looks like it to a test that
only asks "is there excess mass at $\Omega_G$?". `seas_test()`'s per-harmonic
table (`spec$harmonic_table`) and its derived `comb_note` exist to catch
this: they check whether the excess is elevated at *every* seasonal harmonic
(a genuine comb) or only a strict subset (consistent with a masquerading
cycle at just that one frequency).

The two worlds above show the genuine-comb signature -- both harmonics
elevated, no note:

```{r comb-genuine}
x_filter <- ssi_examples$x[[which(ssi_examples$world == "filter")]]
seas_test(x_filter)$spec$harmonic_table
seas_test(x_filter)$spec$comb_note
```

Contrast a constructed pure AR(2) cycle at $\pi/2$ (quarterly annual
frequency), $\rho = 0.95$ -- no seasonality at all, just a narrow nonseasonal
peak that happens to sit where seasonality would:

```{r comb-masquerade}
sim_ar2_cycle <- function(rho, ntot, seed) {
  set.seed(seed)
  coefs <- c(0, -rho^2)          # peaks at pi/2 with resultant AR(2) roots
  x <- numeric(ntot); e <- rnorm(ntot)
  for (t in 3:ntot) x[t] <- sum(coefs * x[t - 1:2]) + e[t]
  x[(ntot - 119):ntot]
}
z   <- sim_ar2_cycle(0.95, 520, seed = 1L)
tst <- seas_test(z, frequency = 4)
tst$decision
tst$spec$label
tst$spec$harmonic_table
tst$spec$comb_note
```

Here the excess is elevated at $\pi/2$ only, and `comb_note` flags exactly
that -- a signature worth a second look (an AR model fit, a periodogram
zoomed on that one frequency) before trusting the seasonal-adjustment call.

This is also the honest residual behind the guard-based whitener's design
(see `vignette("the-machinery")` and `R/whiten.R`'s documentation for the
full measured matrix): guarding only the harmonic tips (rather than whole
seasonal bins) cuts the false-alarm rate on a $\rho = 0.85$ nonseasonal cycle
from about 0.56 (whole-bin exclusion) to about 0.11-0.15, but a *narrower*
$\rho = 0.95$ cycle -- barely wider than the guard -- still trips detection
roughly 30-44% of the time at $n = 120$. The comb diagnostic doesn't prevent
that false alarm; it gives you a second, cheap signal for catching it after
the fact, precisely in the case just demonstrated.

## Summary

`phase_rule` is not something freqseas can determine from `x` alone, and it
does not pretend otherwise: `seas_ssi()` requires it, `seas_adjust()`'s
default is loudly tagged `[declared identification]` in every print and
summary, and getting it backwards costs real accuracy (as the RMSE table
above shows for the filter world, and as the 30-rep regression confirms for
both). Use `"minimum"` when you believe seasonality distorts the series' own
generating mechanism; use `"zero"` when you believe it is an independent,
additive nuisance; and check the line/comb diagnostics above before trusting
either.
